Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2023-09-01 paper/paper-infer.tex
formatting
final typos

The main aim of this paper is to properly validate the use of LDpred2-auto for inferring multiple genetic parameters.
Here we extend LDpred2-auto and show it is a reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities PIP used for fine-mapping purposes), and $\alpha$ (by extending its model to also include this third parameter).
So, on top of providing competitive PGS, LDpred2-auto can also provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without requiring an independent test set.
An overview of what LDpred2-auto can now provide is presented in Figure \ref{fig:overview}.
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$, on average, in UK Biobank analyses.
We call `HapMap3+' this extended set of 1,444,196 SNPs, and recommend to use it when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.9\textwidth]{overview}}
	\caption{Overview of what LDpred2-auto can now provide. For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large,ding2022polygenic}. HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend to use for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$). CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting). \label{fig:overview}}
\end{figure}

\clearpage
\section*{Results}

@@

\clearpage
\section*{Discussion}

@@

Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
Therefore, we recommend to use the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates. When still obtaining a large $p$ estimate ($>1\%$) with this option, we recommend rerunning LDpred2-auto without this option to get a less conservative estimate (cf.\ simulations).
However, LDpred2-auto estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
We think this is likely due to assortative mating, which causes GWAS effects to be inflated \cite[]{border2022assortative,herzig2023model}, which causes our estimates to be inflated as well. 
Second, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.
Future work could focus on correcting these issues.
Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the new sampling and model), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric (mismatch) between GWAS summary statistics and the LD reference used.
Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by. 
LDpred2-auto can also totally miss some causal variants when they are poorly tagged by the set of variants used.
Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
For now, to identify causal variants, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').

@@

\clearpage
\section*{Materials and Methods}

@@

\clearpage
%\vspace*{3em}

@@

>>>>

The main aim of this paper is to properly validate the use of LDpred2-auto for inferring multiple genetic parameters.
Here we extend LDpred2-auto and show it is a reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities PIP used for fine-mapping purposes), and $\alpha$ (by extending its model to also include this third parameter).
So, on top of providing competitive PGS, LDpred2-auto can also provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without requiring an independent test set.
An overview of what LDpred2-auto can now provide is presented in Figure \ref{fig:overview}.
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$, on average, in UK Biobank analyses.
We call `HapMap3+' this extended set of 1,444,196 SNPs, and recommend using it when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.9\textwidth]{overview}}
	\caption{Overview of what LDpred2-auto can now provide. For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large,ding2022polygenic}. HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend using for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$). CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting). \label{fig:overview}}
\end{figure}

%\clearpage
\section*{Results}

@@

%\clearpage
\section*{Discussion}

@@

Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
Therefore, we recommend using the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates. When still obtaining a large $p$ estimate ($>1\%$) with this option, we recommend rerunning LDpred2-auto without this option to get a less conservative estimate (cf.\ simulations).
However, LDpred2-auto estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
We think this is likely due to assortative mating, which causes GWAS effects to be inflated \cite[]{border2022assortative,herzig2023model}, which causes our estimates to be inflated as well. 
Second, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.
Future work could focus on correcting these issues.
Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the new sampling and model), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric (mismatch) between GWAS summary statistics and the LD reference used.
Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by. 
LDpred2-auto can also miss some causal variants when they are poorly tagged by the set of variants used.
Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
For now, to identify causal variants, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').

@@

%\clearpage
\section*{Materials and Methods}

@@

@@

\vspace*{3em}
%\clearpage
